---
title: "2-14-19"
author: "Nathan Kurtz-Enko"
date: "2/14/2019"
output: pdf_document
---

```{r}
library(tidyverse)
```

## Define a quadratic model function
## Play around with other choices (including linear

```{r}
f<-function(x) 10*x*(1-x)

#f<-function(x) exp(2*x)
#f<-function(x) x+ sin(5*pi*x)
```
##Data points
```{r}
N<-100
```
##number of reps 
```{r}
M<-100
```
##prediction points
```{r}
x0<-.5
```
##noise
```{r}
sig<-1.25
```
## Need this...
```{r}
bldForm<-function(d){
  str<-"y~I(x)"
  while(d>1){
    str<-paste(str,sprintf("I(x^%s)",d),sep="+")
    d<-d-1
  }
  str
}
```
## Generate underlying picture
```{r}
xVals<-seq(0,1,.01)
yModel<-f(xVals)
yActs<-f(x0)+rnorm(M,0,sig)
model.df<-data.frame(x=xVals,y=yModel)
```
##The plot
```{r}
gg1<-ggplot()+
  geom_line(data=model.df,
            aes(x,y),
            size=1,
            color="blue")+
  geom_point(data=NULL,aes(x=x0+.01,y=yActs),
             color="green",size=1)+
  scale_y_continuous(limits=c(-2,6))
gg1
```

##Do some predictions, generate xvals to see model structures
```{r}
xvals<-seq(0,1,by=0.01) ##includes x0
n1<-length(xvals)
```
##Here's where we store all the values
```{r}
predVals<-matrix(nrow=M,ncol=n1)
```
##Build a plot with a model of degree = deg!
```{r}
bldPlot<-function(deg){
  ##repeate M times
  for(m in 1:M){
    ##new training data
    x<-runif(N,0,1)
    y<-f(x)+rnorm(N,0,sig)
    ##build the model
    mod<-lm(formula(bldForm(deg)))
    predVals[m,]<-predict(mod,newdata=data.frame(x=xVals))
  }
  ##reorganize the data to plot it.
  predVals<-t(predVals)
  dim(predVals)<-c(M*n1,1)
  vals.df<-data.frame(x=rep(xvals,M),
                      y=predVals,
                      ##Which model
                      model=factor(rep(1:M,each=n1)))
  ## filter on the prediction point so we can emphasize it.
  valsPred.df <-vals.df %>% 
    filter(x==x0)
 ## add to gg1 
  gg1+
    geom_line(data=vals.df,aes(x,y,group=model),alpha=.1)+
    geom_point(data=valsPred.df,aes(x,y),color="red")+
    ggtitle(sprintf("Model deg=%s",deg))
}
```

##testing
```{r}
## Testing....
bldPlot(3)
```
## Build a sequence of these
```{r}
plts<-list()
maxDeg<-12
for(deg in 1:maxDeg){
  plts[[deg]]<-bldPlot(deg)
}
```
## Testing...
```{r}
plts[[1]]
```

##Use the gridExtra package to combine these plots
```{r}
library(gridExtra)

```
## grid for the grid title
```{r}
library(grid)
```

## Now use grid.arrange
```{r}
grid.arrange(grobs=plts,nrow=2,
             top = textGrob("Bias Variance Trade-off",gp=gpar(fontsize=20,font=3)))
```
## In this sequence, you can see how the variance of the predictions is increasing
## with the degree (=flexibility). However, the bias is clearly decreasing