---
title: "Untitled"
author: "Nathan Kurtz-Enko"
date: "5/2/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
```
#Introduction
At the UCI Machine Learning Repository, there is an interesting data
set combining the geographical locations (longitudes and latitudes) of
music along with numerical audio features extracted from WAV files.
The data and more information can be found at:
http://archive.ics.uci.edu/ml/datasets/Geographical+Original+of+Music

Let's do a preliminary exploration of this data set with an eye
towards developing some lines of research.

Start, as always, with loading tidyverse.
```{r}
suppressMessages(library(tidyverse))

```


Let's read the data. There are two different versions of the
data. One has 68 predictor fields, the other has 116. The larger
includes addtional "chromatic" features. We will use the larger data
set. 

Loading ....
```{r}
data.df <- read.csv("default_plus_chromatic_features_1059_tracks.txt",header=F)

```


Take a peeka nd see what we are looking at
```{r}
head(data.df)
dim(data.df)
```


The predictor fields don't have any relevant descriptors, but we do
want to identify the locations, whicha are in the last two
columns. Pull of the dimensions and add field names for the
longitude and latitude. 
```{r}
(nc <- ncol(data.df))
(nr <- nrow(data.df))
names(data.df)[(nc-1):nc] <- c("lat","lon")

```


As explained at the UCI site, the locations identify country of
origin and the lon/lat values correspond to the locations of the
capitals. To simplify things, we will a dd a site identifier.
```{r}
dataSite.df <- data.df %>%
    group_by(lon, lat)%>%
    summarize(siteTot=n())

head(dataSite.df)

```


How many distinct site??
```{r}
(numSites <- nrow(dataSite.df))
```


Add a site number as a unique identifier of each site
```{r}
dataSite.df$siteNum <- factor(1:numSites)

head(dataSite.df)



```


To aid in working with the data later, it helps to separate the
predictors from the locations.
```{r}
lat.dat <- data.df[,nc-1]
lon.dat <- data.df[,nc]
musicData.dat <- data.matrix(data.df[,1:(nc-2)])


```


Where are these locations. To visualize these, create a data frame
of just the sites. Identify them with numbers (as factors). 
Build a plot the world with the sites indicated by labels.
```{r}
world <- map_data("world")
gg.all <- ggplot()+
    geom_map(data=world, map=world,
                    aes(long, lat, map_id=region),
                    color="white", fill="#7f7f7f", size=0.05,
                    alpha=1/4)+
     geom_label(data=dataSite.df, 
                      aes(lon, lat,label=siteNum),
                      size=1.6, alpha=1)
gg.all



```


Add in region information by hand.
These groupings are open to debate.
```{r}
Latin <- c(1, 2,3)
Africa <- c(4,5,6,17,18,19)
Mediterranean <- c(7,9,10,11,12,14)
MidEast <- c(15,16,20,21)
NorthEurope <- c(8,13)
CenAsia <- c(22,23,24,25)
SEAsia <- c(26,27,28)
EAsia <- c(30,31,32)
Austrailia <- c(29,33)

```


Put the list of sites in together and build the data frame. We need
this so that we can add the region name to our overall dataset
```{r}
vals <-
    c(Latin,Africa,Mediterranean,MidEast,NorthEurope,CenAsia,SEAsia,EAsia,Austrailia)

region.df <- data.frame(region=c(rep("Latin",length(Latin)),
                                 rep("Africa",length(Africa)),
                                 rep("Mediterranean",length(Mediterranean)),
                                 rep("MidEast",length(MidEast)),
                                 rep("NorthEurope",length(NorthEurope)),
                                 rep("CenAsia",length(CenAsia)),
                                 rep("SEAsia",length(SEAsia)),
                                 rep("EAsia",length(EAsia)),
                                 rep("Australia",length(Austrailia))),siteNum=factor(vals))
```


How does this look?
```{r}
region.df
dataSite.df <- dataSite.df %>%
    inner_join(region.df)

names(data.df)
head(dataSite.df)
```


Include the regions in the data frame
```{r}
data.df <- data.df %>%
    inner_join(dataSite.df)

```


Pull of the region information
```{r}
region.dat <- data.df[,nc+2]

```


Take a look at the region and site locations...looks ok
```{r}
library(ggrepel)
gg.all2 <- ggplot()+
    geom_map(data=world, map=world,
                    aes(long, lat, map_id=region),
                    color="white", fill="#7f7f7f", size=0.05,
             alpha=1/4)+
     geom_point(data=dataSite.df, 
                      aes(lon, lat),
                size=1.6, color="blue")+
     geom_label_repel(data=dataSite.df, 
                      aes(lon, lat,label=region),
                size=1.6, alpha=1)
gg.all2


```



#Questions
This is a pretty interesting data set. There is potential for
regression, classification, clustering, and PCA. Play around with it a bit
and come up with a  list of investigations. Try to think of
questions that you can approach using the fairly large collection of
techniques you have at your disposal. 

given more instances of traditional music, is there a process that we can use to classify music as belonging to specific region?

given modern music, is there a process that we can use to determine how influential certain areas' traditional music is?