---
title: "Backwards selection"
author: "Nathan Kurtz-Enko"
date: "3/26/2019"
output: pdf_document
---

##Get started:  Load the libraries
```{r}
library(tidyverse)
library(stringr) ## for string manipulation
library(glmnet)


```


#Buld the data


##County-level election data from 2012
Goal: model county level voter turnout as a function of county
demographics and other factors.

You will need to change this to whereever you keep your data
```{r}
countyOrig.df <- read_csv("~/ADM/class/county_facts.csv")

```


take a peek at the data
```{r}
names(countyOrig.df)
head(countyOrig.df)
summary(countyOrig.df)

```



The field names are a bit cryptic. Use the dictionary of field names.
This makes it easier to read in the original csv file.
```{r}
countyDictionary.df <-
    read.csv("~/ADM/class/county_facts_dictionary.csv")

```


Take a peek...
```{r}
names(countyDictionary.df)
head(countyDictionary.df)


```


County population is PST040214
There is one really small county...
```{r}
with(countyOrig.df,summary(PST045214<10))


```


As is usally the case, we need to clean up the data a bit.
For example, PST040210, PST120214, POP010210 refer to other population estimates.
```{r}
county.df <- countyOrig.df%>%
    ##only keep reasonable sized counties
    filter(PST045214>10)%>%
    ##extra population values
    dplyr::select(-PST040210,-PST120214,-POP010210)%>%
    ##drop out area, we don't care about this
    dplyr::select(-LND110210)%>%
    ##total retail, instead just use per capita value
    dplyr::select(-RTN130207)%>%
    dplyr::select(-area_name) %>% 
    ##Change other total fields to  per capita
    mutate(BZA010213=BZA010213/PST045214,
           BZA110213=BZA110213/PST045214,
           NES010213=NES010213/PST045214,
           SBO001207=SBO001207/PST045214,
           MAN450207=MAN450207/PST045214,
           WTN220207=WTN220207/PST045214,
           AFN120207=AFN120207/PST045214,
           BPS030214=BPS030214/PST045214)


```


Take a peek....we still have 40 fields
```{r}
dim(county.df)
summary(county.df)
head(county.df)
names(county.df)

```


Suppose you want to look up a field description real fast.
Here's one way to look at field descriptions.
```{r}
countyDictionary.df%>%
    filter(column_name=="POP060210")


```


Just to get oriented, 
build a linear  model on everything...select out the descriptor fields

Just for fun.....
```{r}
mod <- lm(POP060210~.,data=county.df)

```


Model summary
```{r}
summary(mod)


```



##County level election results
Add in county-level election results from 2012
```{r}
electOrig.df <-
    read.csv("~/ADM/class/US_elect_county.csv")

head(electOrig.df)


```


Clean up and mutate
```{r}
elect.df <- electOrig.df %>%
    filter(fips>0)%>%
    ##zeros in denominators
    filter(ObamaPerc!="#DIV/0!",
           RomneyPerc!="#DIV/0!")%>%
    ##use stringr functions to clean up numbers
    mutate(ObamaVote=str_replace_all(ObamaVote,",",""),
           RomneyVote=str_replace_all(RomneyVote,",",""))%>%
    ##convert to numbers
    mutate(ObamaVote=as.numeric(as.character(ObamaVote)),
           RomneyVote=as.numeric(as.character(RomneyVote)),
           ObamaPerc=as.numeric(as.character(ObamaPerc)),
           RomneyPerc=as.numeric(as.character(RomneyPerc)),
           TotalVote=RomneyVote+ObamaVote)
head(elect.df)


```


Who won???
```{r}
with(elect.df,sum(ObamaVote)/sum(TotalVote))
with(elect.df,sum(RomneyVote)/sum(TotalVote))



```



##Join county and election data
Now we combine the county demographics with the election
results. We also clean up some more fields.
This gives county level demographics and election outcomes.
Only keep total vote values.
```{r}
names(county.df)
names(elect.df)

countyElect.df <- county.df%>%
    ##right join to ensure we get all the counties
    right_join(elect.df)%>%
    ##get rid of these values
    filter(state_abbreviation !="",
           !is.na(ObamaPerc))%>%
    ##don't need these fields
    dplyr::select(-fips,-state_abbreviation,-State,-County,-ObamaVote,-RomneyVote,-ObamaPerc,-RomneyPerc)%>%
    mutate(VotePerc=TotalVote/PST045214)

head(countyElect.df)
names(countyElect.df)
summary(countyElect.df)


```


The whole shebang model
```{r}
mod <- lm(VotePerc~.,data=countyElect.df)
    

```


what do we have??
```{r}
summary(mod)

```


Lots of variables look interesting..which ones to we keep

Pull off the p-values
```{r}
pvals <- summary(mod)$coefficients[,4]
```


grab the statistically significant values
```{r}
sigPVals <- pvals<0.05

sigFields <- names(countyElect.df)[sigPVals]
sigFields

```


take a look at these...
```{r}
sigDescr <- countyDictionary.df%>%
    filter(column_name %in% sigFields)
```


and here they are
```{r}
sigDescr


```

```{r}
sampleSize <- nrow(countyElect.df)

```

```{r}
mseCV <- function(data.df,kfolds=10){
    folds <- sample(1:kfolds,sampleSize,rep=T)
    mse <- rep(0,kfolds)
    for(k in 1:kfolds){
        train.df <- data.df[folds !=k,]
        test.df <- data.df[folds==k,]
        mod <- lm(VotePerc~.,data=train.df)
        vals <- predict(mod,newdata=test.df)
        mse[k] <- with(test.df,mean((VotePerc-vals)^2))
    }
    mean(mse)
}

```

```{r}
#num predictors
numPreds <- length(names(countyElect.df))-1

#predictor indices
allPreds <- 1:(numPreds)

#currently used predictors
currPreds <- c()

#available or non-used predictors
availPreds <- setdiff(allPreds,currPreds)

#max num predictors
maxPreds <- numPreds

#min mse's computed during ensuing loops
minMSE <- numeric(maxPreds)
```

```{r}
stuff <- tibble(Preds = 1:(numPreds-1), mse = 1:(numPreds-1), num = 1:(numPreds-1))

tot <- maxPreds
while(tot>0){
  allMSE <- numeric(length(availPreds))
  ctr <- 1
  for(id in availPreds){
    data <- countyElect.df[,c(availPreds[-id], numPreds+1)]
    mse <- mseCV(data)
    allMSE[ctr] <- mse
    ctr <- ctr+1
  }
  id <- which.min(allMSE)
  id
  worstPred <- availPreds[id]
  worstPred
  bestPreds <- availPreds[-id]
  bestPreds
  bestMSE <- min(allMSE)
  stuff[tot-1,1] = paste(bestPreds, collapse = ",")
  stuff[tot-1,2] = bestMSE
  currPreds <- c(currPreds,worstPred)
  currPreds
  tot <- tot-1
  #maxMSE[tot] <- worstMSE
  availPreds <- setdiff(allPreds,currPreds)
  availPreds
  #availPreds
  ## Print stuff out for debugging and attention-grabbing
  #print(sprintf("Predictor removed: %s  MSE Value: %s",worstPred,worstMSE))
  #print(currPreds)
}

stuff <- add_row(stuff, Preds = paste(allPreds, collapse = ","), mse = mseCV(countyElect.df), num = numPreds)
```

```{r}
minstuff <- stuff%>%
  filter(mse == min(stuff$mse))
minPreds <- minstuff$Preds
minNumPreds <- minstuff$num
```

```{r}
ggplot(stuff)+
  geom_point(aes(x = num, y = mse))+
  geom_line(aes(x = num, y = mse))+
  ggtitle("Backwards Selection on countyElect.df")+
  xlab("Number of Predictors Used")+
  ylab("MSE")
```

The MSE begins to level off at around 35 predictors using backward selection.