---
title: "Homework 18"
author: "Nathan Kurtz-Enko"
date: "4/25/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
suppressMessages(library(tidyverse))
suppressMessages(library(factoextra))
suppressMessages(library(ggrepel))
suppressMessages(library(MASS))
library(cluster)
```

#Assignment: Clustering on county_facts.csv

```{r}
county_facts.df <- read_csv("/home/rstudio/users/kurtze1/ADM/class/county_facts.csv")

#clean data
county_facts.df <- county_facts.df%>%
    ##only keep reasonable sized counties
    filter(PST045214>10)%>%
    ##extra population values
    dplyr::select(-PST040210,-PST120214,-POP010210)%>%
    ##drop out area, we don't care about this
    dplyr::select(-LND110210)%>%
    ##total retail, instead just use per capita value
    dplyr::select(-RTN130207)%>%
    dplyr::select(-area_name) %>% 
    ##Change other total fields to  per capita
    mutate(BZA010213=BZA010213/PST045214,
           BZA110213=BZA110213/PST045214,
           NES010213=NES010213/PST045214,
           SBO001207=SBO001207/PST045214,
           MAN450207=MAN450207/PST045214,
           WTN220207=WTN220207/PST045214,
           AFN120207=AFN120207/PST045214,
           BPS030214=BPS030214/PST045214)

#remove NA values
county_facts.df <- na.omit(county_facts.df)

#scale data
county_facts_stuff <- scale(county_facts.df[, 3:ncol(county_facts.df)])
county_facts.df <- cbind(county_facts.df[1:2], county_facts_stuff)
```


```{r, warning=FALSE}
K <- 3 
mod.km <- kmeans(county_facts_stuff,K,nstart=25)
county_facts.df$cluster <- factor(mod.km$cluster)

fviz_cluster(mod.km,data=county_facts_stuff)

sil <- silhouette(mod.km$cluster,get_dist(county_facts_stuff))

avg_si <- function(k) {
  mod.km<- kmeans(county_facts_stuff, centers = k, nstart = 25)
  ss <- silhouette(mod.km$cluster, get_dist(county_facts_stuff))
  mean(ss[, 3])
}

k.values <- 2:15


avg_si_vals <- map_dbl(k.values, avg_si)

data.frame(k=k.values,si=avg_si_vals)%>%
    ggplot()+
    geom_point(aes(k,si))+
    geom_line(aes(k,si))+
    scale_x_continuous(breaks=k.values)+
    ggtitle("Average silhoutte values")
```


#Assigment: Prediction with Eigenfaces
Does prediction into the dog or cat eigenface reliability
predict if an images is a dog or a cat? That is, for faces not in
the training sets, does the quality of the reconstruction indicate
"catness" versus "dogness"?

Plan: for a dog (or cat) reconstruct with both the eigenCats and eigenDogs. Which is "closer" to the original. Does closeness predict the category (cat vs dog) well?

Things to consider

 * What is the minimal number of eigenfaces needed to reliability
predict cat vs dog? Extreme case: can you use the just first eigenface?
 * In the dog set, which dog is most like a cat? Which dog is the
least like a cat? How about cats like/unlike dogs?
 * In the human face is it more like a cat or a dog? Do the whiskers change things?
 
 * Use your own face image. Are you more like a cat or a dog? Note: Use *readJPG.R* to convert a jpeg image to a form suitable for this process.
 
```{r}
catsAll <- read.csv("~/ADM/class/cats.csv",header=F)
dogsAll <- read.csv("~/ADM/class/dogs.csv",header=F)
facesAll <- read.csv("~/ADM/class/face.csv",header=F)

library(jpeg)

fig <- readJPEG("/home/rstudio/users/kurtze1/ADM/homework/IMG_1326.jpg")
mpr <- t(255*fig[,,1])
image(mpr,col=grey.colors(256))
mprFlat <- matrix(t(mpr),nrow=64^2)
write.csv(mprFlat,"mpr.csv",row.names=F)

```

```{r}
cats <- catsAll[,1:90]
catsOther <- catsAll[,91:99]
dogs <- dogsAll[,1:90]
dogsOther <- dogsAll[,91:99]
```

```{r}
imageConv <- function(inCat){
    matrix(inCat,nrow=64,byrow=T)    
}
flatConv <- function(inCat){
    matrix(t(inCat),nrow=64^2,ncol=1)[,1]
}
```

```{r}
meanCats <- colMeans(cats)
meanDogs <- colMeans(dogs)

meanCatMat <- matrix(rep(meanCats,each=4096),nrow=4096)
meanDogMat <- matrix(rep(meanDogs,each=4096),nrow=4096)
```
 
```{r}
cat.svd <- svd(cats)
dog.svd <- svd(dogs)
```

```{r}
U.cat <- cat.svd$u
D.cat <- cat.svd$d
V.cat <- cat.svd$v

U.dog <- dog.svd$u
D.dog <- dog.svd$d
V.dog <- dog.svd$v
```
 
```{r}
CatCat <- function(){
  num <- 5
  samp <- sample(1:9,num,rep=F)
  catHats <- matrix(nrow=4096,ncol=num)
  for(i in 1:num){
      catHats[,i] <- U.cat%*%t(U.cat)%*%catsOther[,i]
  }
  newOrd <- sample(1:num,num,rep=F)
  op <- par(mfrow=c(2,num))
  for(i in 1:num){
      image(imageConv(catsOther[,i]),col=grey.colors(256))
  }
  for(i in 1:num){
      image(imageConv(catHats[,newOrd[i]]),col=grey.colors(256))
  }
  par(op)
}


DogDog <- function(){
  num <- 5
  samp <- sample(1:9,num,rep=F)
  dogHats <- matrix(nrow=4096,ncol=num)
  for(i in 1:num){
      dogHats[,i] <- U.dog%*%t(U.dog)%*%dogsOther[,i]
  }
  newOrd <- sample(1:num,num,rep=F)
  op <- par(mfrow=c(2,num))
  for(i in 1:num){
      image(imageConv(dogsOther[,i]),col=grey.colors(256))
  }
  for(i in 1:num){
      image(imageConv(dogHats[,newOrd[i]]),col=grey.colors(256))
  }
  par(op)
}


DogCat <- function(){
  num <- 5
  samp <- sample(1:9,num,rep=F)
  dogcatHats <- matrix(nrow=4096,ncol=num)
  for(i in 1:num){
      dogcatHats[,i] <- U.dog%*%t(U.dog)%*%catsOther[,i]
  }
  newOrd <- sample(1:num,num,rep=F)
  op <- par(mfrow=c(2,num))
  for(i in 1:num){
      image(imageConv(catsOther[,i]),col=grey.colors(256))
  }
  for(i in 1:num){
      image(imageConv(dogcatHats[,newOrd[i]]),col=grey.colors(256))
  }
  par(op)
}

CatDog <- function(){
  num <- 5
  samp <- sample(1:9,num,rep=F)
  catdogHats <- matrix(nrow=4096,ncol=num)
  for(i in 1:num){
      catdogHats[,i] <- U.cat%*%t(U.cat)%*%dogsOther[,i]
  }
  newOrd <- sample(1:num,num,rep=F)
  op <- par(mfrow=c(2,num))
  for(i in 1:num){
      image(imageConv(dogsOther[,i]),col=grey.colors(256))
  }
  for(i in 1:num){
      image(imageConv(catdogHats[,newOrd[i]]),col=grey.colors(256))
  }
  par(op)
}
```
 
```{r}
CatCat()
DogCat()

DogDog()
CatDog()
```
 


 Both the DogCat() and CatDog() functions do an awful job of predicting faces. However, if I had to choose, the better of the two would be DogCat(), using dog eigen faces to predict cat faces. But it is far superior to simply use cat eigen faces to predict cat faces and the same goes for dog faces.
 
 
```{r}
#mpr.csv <- read_csv("mpr.csv")
#image(imageConv(as.matrix(mpr.csv)),col=grey.colors(256))
#aDogHat <- U.dog%*%t(U.dog)%*%as.matrix(mpr.csv)
#image(imageConv(aDogHat),col=grey.colors(256))

#aCatHat <- U.cat%*%t(U.cat)%*%as.matrix(mpr.csv)
#image(imageConv(aCatHat), col = grey.colors(256))
```
 
 