---
title: "homework_19.Rmd"
author: "Nathan Kurtz-Enko"
date: "4/30/2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggdendro)
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(ggrepel))
suppressMessages(library(graphics))
suppressMessages(library(factoextra))
library(wbstats)
```

## Assignment

### Find another data set similar to the NBA dataset
(Baseball Reference is one choice, but there are other, non-sport
options). Perform a complete cluster and PCA analysis.

```{r}
#data

data <- read_csv("/home/rstudio/users/kurtze1/Mscs 264 F18/Project/Shaoze and Nathan /soccer_data.csv")

data <- filter(data, points > 50)

#principal components

club <- data$club

other <- dplyr::select(data, points, wins, draws, losses, goals_for,
                       goals_against, avg_plyr_salary, yellow_card, 
                       red_card, passing, possession, rating)
other.scaled <- scale(other)

#add row names
row.names(other.scaled) <- club
#pca club and league

club.pca <- prcomp(other.scaled)


#plots
fviz_pca_biplot(club.pca,repel=TRUE)

#K MEANS
K <- 5
club.km <- kmeans(other.scaled,K,iter.max=20,nstart=30)

#the clusters

data <- mutate(data, cluster = club.km$cluster)

#pca by hand
rots <- club.pca$rotation
club.pca1<- data.matrix(other.scaled) %*% rots
pca1 <- club.pca1[,1]
pca2 <- club.pca1[,2]
pca3 <- club.pca1[,3]

#build new data frame for pretty plot
clubCluster.df <- data.frame(pca1,pca2,cluster=factor(data$cluster),name=club)


#new pretty plot
ggplot(clubCluster.df,aes(pca1,pca2,color=cluster))+
    geom_point(size=1)+
    geom_text_repel(aes(label=name),size=3)+
    ggtitle("Club Cluster Analysis and PCA")

#heirarch clust
club.dist <- dist(other.scaled)

#complete
club.hc.c <- hclust(club.dist,method="complete")

#plot
par(mfrow=c(1,1))
plot(club.hc.c,cex=0.5,main="Complete")
```

##Assignment: World Bank Data
Extract a more interesting set of parameters. Explore the World
Bank site and get at least 12 different (more is ok)
features. Select at least two different years (separated in
time). How do the clusters looks? If you add Principal Components,
do you see any natural clustering behavior relative to the
Principal Components?

```{r}
new_wb_cache <- wbcache()
```

```{r}
elec_access <- "EG.ELC.ACCS.ZS"
farm_land <- "AG.LND.AGRI.ZS"
fresh_water_use <- "ER.H2O.FWTL.K3"
c02_emmiss <- "EN.ATM.CO2E.KT"
contraceptive_prev <- "SP.DYN.CONU.ZS"
birth_rt <- "SP.DYN.CBRT.IN"
death_rt <- "SP.DYN.CDRT.IN"
immuinza_perc <- "SH.IMM.IDPT"
life_exp <- "SP.DYN.LE00.IN"
gdp_per_cap_growth <- "NY.GDP.PCAP.KD.ZG"
gdp_per_cap <- "NY.GDP.PCAP.CD"
gdp_growth <- "NY.GDP.MKTP.KD.ZG"
savings_as_perc_gdp <- "NY.GNS.ICTR.ZS"
inflation <- "FP.CPI.TOTL.ZG"

inds <- c(elec_access, farm_land, c02_emmiss, birth_rt, death_rt, immuinza_perc, 
          life_exp, gdp_per_cap_growth, gdp_per_cap, gdp_growth, 
          savings_as_perc_gdp, inflation)

wb_dat <- wb(indicator = inds)

wb2014 <- wb_dat %>%
    filter(date=="2014")

wb2000 <- wb_dat %>%
  filter(date == "2000")


wb_countries <- wbcountries() 
allCountries <- wb_countries %>%
    filter(region != "Aggregates") %>%
    dplyr::select(country)

wb2014 <- wb2014 %>% 
    filter(country %in% allCountries$country) %>%
    dplyr::select(-iso3c,-iso2c,-date,-indicator)
wb2014x <- wb2014 %>%
    spread(indicatorID,value)

wb2000 <- wb2000 %>% 
    filter(country %in% allCountries$country) %>%
    dplyr::select(-iso3c,-iso2c,-date,-indicator)
wb2000x <- wb2000 %>%
    spread(indicatorID,value)

cc2014 <- complete.cases(wb2014x)
cc2000 <- complete.cases(wb2000x)
wb2014.df <- wb2014x[cc2014,] %>% filter()
wb2000.df <- wb2000x[cc2000,]
```

```{r}
#pca 2014 

country <- wb2014.df$country

other <- wb2014.df[, -1]
other.scaled <- scale(other)
row.names(other.scaled) <- country
row.names(other) <- country
pca.2014 <- prcomp(other)

fviz_pca_biplot(pca.2014,repel=TRUE)

#K MEANS
K <- 5
club.km <- kmeans(other.scaled,K,iter.max=20,nstart=30)

#the clusters

data <- mutate(data, cluster = club.km$cluster)

#pca by hand
rots <- club.pca$rotation
club.pca1<- data.matrix(other.scaled) %*% rots
pca1 <- club.pca1[,1]
pca2 <- club.pca1[,2]
pca3 <- club.pca1[,3]

#build new data frame for pretty plot
clubCluster.df <- data.frame(pca1,pca2,cluster=factor(data$cluster),name=club)


#new pretty plot
ggplot(clubCluster.df,aes(pca1,pca2,color=cluster))+
    geom_point(size=1)+
    geom_text_repel(aes(label=name),size=3)+
    ggtitle("Club Cluster Analysis and PCA")

#heirarch clust
club.dist <- dist(other.scaled)

#complete
club.hc.c <- hclust(club.dist,method="complete")

#plot
par(mfrow=c(1,1))
plot(club.hc.c,cex=0.5,main="Complete")
```

```{r}
#pca 2000

country <- wb2000.df$country

other <- wb2000.df[, -1]
other.scaled <- scale(other)
row.names(other.scaled) <- country
row.names(other) <- country
pca.2000 <- prcomp(other)

fviz_pca_biplot(pca.2000,repel=TRUE)

#K MEANS
K <- 5
club.km <- kmeans(other.scaled,K,iter.max=20,nstart=30)

#the clusters

data <- mutate(data, cluster = club.km$cluster)

#pca by hand
rots <- club.pca$rotation
club.pca1<- data.matrix(other.scaled) %*% rots
pca1 <- club.pca1[,1]
pca2 <- club.pca1[,2]
pca3 <- club.pca1[,3]

#build new data frame for pretty plot
clubCluster.df <- data.frame(pca1,pca2,cluster=factor(data$cluster),name=club)


#new pretty plot
ggplot(clubCluster.df,aes(pca1,pca2,color=cluster))+
    geom_point(size=1)+
    geom_text_repel(aes(label=name),size=3)+
    ggtitle("Club Cluster Analysis and PCA")

#heirarch clust
club.dist <- dist(other.scaled)

#complete
club.hc.c <- hclust(club.dist,method="complete")

#plot
par(mfrow=c(1,1))
plot(club.hc.c,cex=0.5,main="Complete")
```

