---
title: "EDA_2019_01_29"
author: "Andrew Srisuwananukorn"
date: "2/2/2019"
output: html_document
---
  
#Introduction
  - exploration of cheese data provided by Thomas Mock of Tidy Tuesday
https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-01-29

- Also inspired by clustering methods DataCamp lesson by Dmitry 

Modified for MSCS  341: Algorithms for Decision Making

# Standard startup.
```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Import libraries
```{r}
library(tidyverse)
library(ggplot2)
library(purrr)
library(cluster)
library(dendextend)

```


#Import Data...take a look after done.


Set up the data import. Make sure you have a well-organized structure for storing your code and data for this course!
```{r}
fileName <- "clean_cheese.csv"
dataDir <- "~/ADM/"
dataPath <- file.path(dataDir,fileName)
```

The values are for cheese consumption by cheese type.
```{r,message=TRUE}

raw_cheese.df <- read_csv(dataPath) 

```
## Reorganize the data
The data are organized in a fairly standard form. However, for analysis, we need to some reshaping and filters.

Different versions of the data

## Version 1: gather all the values from columns after the Year and assign the column name as the key
```{r}
cheese1.df <- raw_cheese.df %>% 
  gather(key = Cheese, value = value, 2:ncol(raw_cheese.df))
```


## Version 2: Separate out the "Totals" 
```{r}
cheeseTot.df<-cheese1.df %>% 
  filter(str_detect(Cheese,"Total"))

cheeseNoTot.df<-cheese1.df %>% 
  filter(!str_detect(Cheese,"Total"))
```



# Version 3: Spread the Year out. (Note: spread and gather are "inverses")
```{r}

cheese.df <- cheeseNoTot.df %>% 
  spread(key = "Year", value = "value") 

```



# Plot the data

## First just the raw data. This makes it hard to compare across cheese types
```{r}
raw_cheese.df %>% 
  ggplot(aes(x = Year, y = Cheddar))+
  geom_point()+
  geom_line() +
  geom_smooth(method = "loess")

```

To add another cheese, we need another stanza. Awkward
```{r}
raw_cheese.df %>% 
  ggplot()+
  geom_point(aes(Year,Cheddar))+
  geom_line(aes(Year,Cheddar))+
  geom_point(aes(Year,Swiss))+
  geom_line(aes(Year,Swiss))
  
```


Much better with cheese1.df. This is because each Year+Cheese is an observation
```{r}
cheese1.df %>% 
  ggplot(aes(Year,value,color=Cheese))+
  geom_point(size=.2, na.rm = TRUE)+
  geom_line(na.rm = TRUE)
```

Plot just the totals
```{r}
cheeseTot.df %>% 
  ggplot(aes(Year,value,color=Cheese))+
  geom_point(size=.2)+
  geom_line()
```

Plot without the totals
```{r}
cheeseNoTot.df %>% 
  ggplot(aes(Year,value,color=Cheese))+
  geom_point(size=.2)+
  geom_line()
```

#Scale data
This is unnecessary for this specific data frame, as the scale is the same for each feature (=Year).


To make this easier, move the first column to the row name.
Then scale. Note: this makes the mean/var of each column (Year) 0/1. 
```{r}
scaled_cheese.df <- cheese.df %>% 
  column_to_rownames("Cheese") %>% 
  scale() 

  
```
Compute the (euclidian) distance matrix, that is how far apart are the cheeses?
```{r}
scaled_dist_cheese <- scaled_cheese.df%>% 
  dist()

```
To see this as a matrix of distances, convert.

Note that Blue is close to Brick, but not to American Other.
```{r}
round(as.matrix(scaled_dist_cheese),2)
```



#Hierarchical clustering
Now we can cluster to see how close one cheese is to another.

We will use Hierarchical clustering, a topic we'll cover later in the semester.

Parameters for hierarchical clustering.
Distance = "euclidian"
hierarchical clustering method = "complete"

```{r}
hc_cheese <- hclust(scaled_dist_cheese, method = "complete")
```

## First look
```{r}
hc_cheese %>% as.dendrogram() %>% 
  set("branches_k_color") %>% 
  plot()
```


#By visual inspection, optimal k=2 (= number of clusters)
```{r}
par(mar = c(5,4,1,6)) #Reset margins: bottom, left, top, right
hc_cheese %>% as.dendrogram() %>% 
  set("branches_k_color", k = 2) %>% 
  set("labels_cex", 0.75) %>% 
  plot(horiz = TRUE,
       main = "Hierarchical Clustering by Cheese Consumption", 
       sub = "k=2 by visual inspection") %>% 
  abline(v = 15, lty = 2)
```



#Assign cluster groups
```{r}
clusters_hc <- cutree(hc_cheese, k = 2)
```

##Plot dendogram using ggplot tools
```{r}
library(ggdendro)
ggdendrogram(hc_cheese,rotate=TRUE)+
  ggtitle("Clusters of Cheese")

```
# Growth rate analysis

Instead of looking at totals, let's cluster on the rate of growth of cheese production. To compute the differences (rates), use the "lag" function of dplyr.


Repeat everything above
```{r}
cheeseRate.df<-cheeseNoTot.df %>% 
  ##get rid of Blue since it has annoying NAs
  filter(Cheese != "Blue") %>% 
  ##group and use lag
  group_by(Cheese) %>% 
  mutate(value2=lag(value),
         rate=(value-value2)/value) %>% 
  filter(!is.na(value2)) %>% 
  select(-value,-value2)
names(cheeseRate.df)
```
Plot

```{r}
cheeseRate.df %>% 
  #filter(Cheese=="Cheddar") %>% 
  ggplot(aes(Year,rate,color=Cheese))+
  geom_point(size=.5)+
  geom_line()+
  ggtitle("Grow rates of Cheese Comsumption")
```

Doesn't look like there is much going on. Just for fun, repeat the steps from above with the new data frame.

```{r}
cheeseRate2.df <- cheeseRate.df %>% 
  spread(key = "Year", value = "rate")

scaled_cheeseRate.df <- cheeseRate2.df %>% 
  column_to_rownames("Cheese") %>% 
  scale() 

scaled_dist_cheeseRate <- scaled_cheeseRate.df %>% 
  dist()
```




```{r}
hc_cheeseRate <- hclust(scaled_dist_cheeseRate, method = "complete")
```

## Things look different now.

It's not as clear how to cluster with these values.
```{r}
hc_cheeseRate %>% as.dendrogram() %>% 
  set("branches_k_color") %>% 
  plot()
```



Take a look with the ggplot version.


```{r}
ggdendrogram(hc_cheeseRate,rotate=TRUE)+
  ggtitle("Clusters of Cheese")
```

#K-means analysis

We will use the elbow plot to determine optimal k
 Use map_dbl to run many models with varying value of k (centers)


```{r}
tot_withinss <- map_dbl(1:10,  function(k){
  model <- kmeans(x = scaled_dist_cheese, centers = k)
  model$tot.withinss
})
tot_withinss
```

# Generate a data frame containing both k and tot_withinss
```{r}
elbow.df <- data.frame(
  k = 1:10 ,
  tot_withinss = tot_withinss
)
ggplot(elbow.df, aes(x = k , y = tot_withinss)) +
  geom_line() +
  scale_x_continuous(breaks = 1:10) + 
  labs(title = "Elbow Plot for K-means clustering of Cheese Consumption",
       subtitle = "Optimal k=2 by visual inspection of elbow point",
       y = "Total within squared sums",
       x = "k") +
  geom_point(data = elbow.df %>% filter(k == 2), aes(x = k, y = tot_withinss, color = "red", size = 3), show.legend = FALSE)

#Optimal k = 2
km_cheese <- kmeans(scaled_dist_cheese, centers = 2)
clusters_km <- km_cheese$cluster


```
Arguably, k=3 is ok too.

#Partioning Around Mediods (PAM) and silhouette
- pam is from the cluster pacakage
```{r}
#using the silhouette technique to determine optimal k

# Use map_dbl to run many models with varying value of k
sil_width <- map_dbl(2:10,  function(k){
  model <- pam(x = scaled_dist_cheese, k = k)
  model$silinfo$avg.width
})

# Generate a data frame containing both k and sil_width
sil.df <- data.frame(
  k = 2:10,
  sil_width = sil_width
)

# Plot the relationship between k and sil_width
ggplot(sil.df, aes(x = k, y = sil_width)) +
  geom_line() +
  scale_x_continuous(breaks = 2:10) +   
  labs(title = "Silhouette Plot for K-means clustering of Cheese Consumption",
       subtitle = "Optimal k=2 by maximal silhouette width",
       y = "Average silhouette width",
       x = "k") +
  geom_point(data = sil.df %>% filter(k == 2), aes(x = k, y = sil_width, color = "red", size = 3), show.legend = FALSE)
```


#Optimal k = 2

```{r}

pam_cheese <- pam(scaled_dist_cheese, k = 2)
clusters_pam <- pam_cheese$clustering

```

#Put together a plot of the  consumptions totals, grouped by clusters.

```{r,message=F,warning=F}
total_cheese <- mutate(cheese.df, 
                       cluster_hc = clusters_hc,
                       cluster_km = clusters_km,
                       cluster_pam = clusters_pam) %>%
  gather(-c("Cheese","cluster_km", "cluster_hc", "cluster_pam"),
                              key = "year", value = "eaten") %>% 
  ##so we can change the cluster method easily                            
  mutate(cluster=cluster_hc)
method<-"HC"
```

And now the plot.

```{r}
total_cheese %>% 
  ggplot(aes(x = year, y = eaten, color = factor(cluster))) + 
  geom_line(aes(group = Cheese),show.legend = FALSE) + 
  geom_smooth(aes(group = cluster),se=FALSE, na.rm=TRUE, show.legend = FALSE) +
  geom_text(data = total_cheese %>% filter(year == 2013), 
            aes(label = Cheese), size = 2.5, nudge_y = 0.5, show.legend = FALSE) +
  scale_x_discrete(breaks = seq(1970,2015, 5)) +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title = "Average cheese consumption from USDA",
       subtitle = sprintf("Optimal k=2 by Elbow method of %s cluster analysis",method),
       y = "Average cheese consumption (lbs per person)",
       x = "Year"
  )
```
Alternative: use ggrepel package to place the labels better.
```{r}
library(ggrepel)
total_cheese %>% 
  ggplot(aes(x = year, y = eaten, color = factor(cluster))) + 
  geom_line(aes(group = Cheese),show.legend = FALSE) + 
  geom_smooth(aes(group = cluster),se=FALSE, na.rm=TRUE, show.legend = FALSE) +
  geom_label_repel(data = total_cheese %>% filter(year == 2013), 
            aes(label = Cheese), size = 2.5,  show.legend = FALSE,
            color="black") +
  scale_x_discrete(breaks = seq(1970,2015, 5)) +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title = "Average cheese consumption from USDA",
       subtitle = sprintf("Optimal k=2 by Elbow method of %s cluster analysis",method),
       y = "Average cheese consumption (lbs per person)",
       x = "Year"
  )
```