---
title: "4-25-19"
author: "Nathan Kurtz-Enko"
date: "4/25/2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
```
Load libraries
```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(factoextra))
suppressMessages(library(ggrepel))



```


Image data
```{r}
catsAll <- read.csv("cats.csv",header=F)
dogsAll <- read.csv("dogs.csv",header=F)
facesAll <- read.csv("face.csv",header=F)



```


Each dataset cats and dogs contain 99 images of cats and dogs. Each is 64x64
image stacked into a row  4096=64*64 greyscale values
```{r}
dim(catsAll)
dim(dogsAll)

```


Subset this into  training and testing sets
```{r}
cats <- catsAll[,1:90]
catsOther <- catsAll[,91:99]
dogs <- dogsAll[,1:90]
dogsOther <- dogsAll[,91:99]

```


Extract the dimensions
```{r}
(n <- nrow(cats))
(p <- ncol(cats))
```


Note that n= number of observations, an observation  is a pixel
value.  Also, p = number of features, in this case, its a the pixal
value across all the cats!


facesAll is a pair of faces, with the same characteristics
```{r}
dim(facesAll)


```



##Helper functions

This converts an image into 64x64
```{r}
imageConv <- function(inCat){
    matrix(inCat,nrow=64,byrow=T)    
}

```


This packs into a 64*64=4096 list
```{r}
flatConv <- function(inCat){
    matrix(t(inCat),nrow=64*64,ncol=1)[,1]
}


```


##Looking at images
Here's how to plot an image
```{r}
cat1 <- cats[,1]
image(imageConv(cat1),col=grey.colors(256))

```


Look at 9 randomly selected cats
```{r}
samp <- sample(1:p,9,rep=F)
catSamp <- cats[,samp]
op <- par(mfrow=c(3,3))
for(i in 1:9){
    image(imageConv(catSamp[,i]),col=grey.colors(256))
}
par(op)


```


same for dogs....
Look at 9 randomly selected cats
```{r}
samp <- sample(1:p,9,rep=F)
dogSamp <- dogs[,samp]
op <- par(mfrow=c(3,3))
for(i in 1:9){
    image(imageConv(dogSamp[,i]),col=grey.colors(256))
}
par(op)


```


##Singular Value Decomposition
Now we can begin the computation in the image space
To make the SVD work, we need to center the data.
```{r}
(meanCats <- colMeans(cats))
```


subtract of the means
```{r}
meanCatMat <- matrix(rep(meanCats,each=4096),nrow=4096)
```


Subtract and check
```{r}
cats0 <- cats-meanCatMat
```


Check...should be zero
```{r}
colMeans(cats0)


```


Now the Singular Value Decomposition
```{r}
cat.svd <- svd(cats)


```


Here are the u, d, and v matrices
```{r}
U <- cat.svd$u
D <- cat.svd$d
V <- cat.svd$v

```


Check the dimensions
```{r}
dim(U)
length(D)
dim(V)

```


And properties...U and V areorthonormal
U
```{r}
t(U)%*% U
```


V
```{r}
t(V)%*% V


```


The diagonal values d Measuresthe variabililty contained in each component
Pack into data frame
```{r}
proj.df <- data.frame(rank=1:p,var=D,totvar=cumsum(D)/sum(D)) 

```


What do we have
```{r}
proj.df %>%
    gather(type,val,var:totvar) %>% 
    ggplot()+
    geom_point(aes(rank,val))+
        geom_line(aes(rank,val),color="blue")+
    facet_wrap(~type,scales="free")+
    ggtitle("Variablity by rank")
```


From this we see that the accumulation of variation is pretty slow,
no reasonably small set of faces will reconstruct well. The first
face does contain 20% of the variability.



#Eigenface reconstructions
The columns of the U matrix are the eigenfaces
The first eigenface
```{r}
eigenFace1=U[,1]

```


We can use the "Hat" approach: yhat=UU'Y
```{r}
par(mfrow=c(1,1))
image(imageConv(eigenFace1),col=grey.colors(256))

```


Look at first 9 eigenfaces
```{r}
samp <- 1:9
catSamp <- U[,samp]
op <- par(mfrow=c(3,3))
for(i in 1:9){
    image(imageConv(catSamp[,i]),col=grey.colors(256))
}
par(op)


```


The last 9 eigenfaces
```{r}
samp <- 81:90
catSamp <- U[,samp]
op <- par(mfrow=c(3,3))
for(i in 1:9){
    image(imageConv(catSamp[,i]),col=grey.colors(256))
}
par(op)



```


Eigenface reconstuctions
Pull out one of the other cats..
```{r}
aCat <- catsOther[,9]
image(imageConv(aCat),col=grey.colors(256))

```


A cat in a hat...
```{r}
aCatHat <- U%*%t(U)%*%aCat
image(imageConv(aCatHat),col=grey.colors(256))

```


Same with shifted image...essentially the same
```{r}
aCat0 <- aCat-mean(aCat)
aCatHat0 <- U%*%t(U)%*%aCat0
image(imageConv(aCatHat0),col=grey.colors(256))

```


Compare together
```{r}
op <- par(mfrow=c(2,2))
image(imageConv(aCat),col=grey.colors(256))
image(imageConv(aCatHat),col=grey.colors(256))
par(op)

```


##Facial Recognition Challange

Reconstruct cat faces and see if we can match original to reconstruct!

The number of images
```{r}
num <- 5
samp <- sample(1:9,num,rep=F)
catHats <- matrix(nrow=4096,ncol=num)
for(i in 1:num){
    catHats[,i] <- U%*%t(U)%*%catsOther[,i]
}
newOrd <- sample(1:num,num,rep=F)
op <- par(mfrow=c(2,num))
for(i in 1:num){
    image(imageConv(catsOther[,i]),col=grey.colors(256))
}
for(i in 1:num){
    image(imageConv(catHats[,newOrd[i]]),col=grey.colors(256))
}
par(op)

```


The answer....
```{r}
op <- par(mfrow=c(2,num))
for(i in 1:num){
    image(imageConv(catsOther[,i]),col=grey.colors(256))
}
for(i in 1:num){
    image(imageConv(catHats[,i]),col=grey.colors(256))
}
par(op)

```


Actual order
```{r}
newOrd
```


## What about a dog??
Pull off a dog!
```{r}
aDog <- dogsAll[,5]
image(imageConv(aDog),col=grey.colors(256))

```


A dog in a hat...Construct a dog out of eigenCats
```{r}
aDogHat <- U%*%t(U)%*%aDog
image(imageConv(aDogHat),col=grey.colors(256))

```


Compare together
```{r}
op <- par(mfrow=c(2,1))
image(imageConv(aDog),col=grey.colors(256))
image(imageConv(aDogHat),col=grey.colors(256))
par(op)



```


###Other faces....
```{r}
aFace <- facesAll[,1]
aFaceHat <- U%*%t(U)%*%aFace

```


Compare
```{r}
op <- par(mfrow=c(2,1))
image(imageConv(aFace),col=grey.colors(256))
image(imageConv(aFaceHat),col=grey.colors(256))
par(op)

```


Other face?
```{r}
aFace <- facesAll[,2]
aFaceHat <- U%*%t(U)%*%aFace

```


Compare
```{r}
op <- par(mfrow=c(2,1))
image(imageConv(aFace),col=grey.colors(256))
image(imageConv(aFaceHat),col=grey.colors(256))
par(op)


```


## Repeat for combined pet faces
What do the eigenfaces look like? Can you distinguish the classes?
```{r}
pets <- cbind(cats,dogs)
(p <- ncol(pets))

```


To make the SVD work, we need to center the data.
```{r}
(meanPets <- colMeans(pets))
```


subtract of the means
```{r}
meanPetMat <- matrix(rep(meanPets,each=4096),nrow=4096)
```


Subtract and check
```{r}
pets0 <- pets-meanPetMat
colMeans(pets0)
```



```{r}
pets.svd <- svd(pets)
U <- pets.svd$u
D <- pets.svd$d
V <- pets.svd$v

```


The diagonal values d Measuresthe variabililty contained in each component
Pack into data frame
```{r}
proj.df <- data.frame(rank=1:p,var=D,totvar=cumsum(D)/sum(D)) 

```


What do we have
```{r}
proj.df %>%
    gather(type,val,var:totvar) %>% 
    ggplot()+
    geom_point(aes(rank,val))+
        geom_line(aes(rank,val),color="blue")+
    facet_wrap(~type,scales="free")+
    ggtitle("Variablity by rank")
```


From this we see that the accumulation of variation is pretty slow,
no reasonably small set of faces will reconstruct well.



New Eigenface
```{r}
image(imageConv(U[,1]),col=grey.colors(256))

```


Look at first 9 eigenfaces
```{r}
samp <- 1:9
petSamp <- U[,samp]
op <- par(mfrow=c(3,3))
for(i in 1:9){
    image(imageConv(petSamp[,i]),col=grey.colors(256))
}
par(op)



```


Convert a dog
```{r}
aDog <- dogsOther[,1]
op <- par(mfrow=c(1,2))
image(imageConv(aDog),col=grey.colors(256))
```


A dog in a hat...
```{r}
aDogHat <- U%*%t(U)%*%aDog
image(imageConv(aDogHat),col=grey.colors(256))
par(op)


```


#Assigment: Prediction with Eigenfaces
Does prediction into the dog or cat eigenface reliability
predict if an images is a dog or a cat? That is, for faces not in
the training sets, does the quality of the reconstruction indicate
"catness" versus "dogness"?

Plan: for a dog (or cat) reconstruct with both the eigenCats and eigenDogs. Which is "closer" to the original. Does closeness predict the category (cat vs dog) well?

Things to consider

 * What is the minimal number of eigenfaces needed to reliability
predict cat vs dog? Extreme case: can you use the just first eigenface?
 * In the dog set, which dog is most like a cat? Which dog is the
least like a cat? How about cats like/unlike dogs?
 * In the human face is it more like a cat or a dog? Do the whiskers change things?
 
 * Use your own face image. Are you more like a cat or a dog? Note: Use *readJPG.R* to convert a jpeg image to a form suitable for this process.


