---
title: "4-30-19"
author: "Nathan Kurtz-Enko"
date: "4/30/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
```
#Introduction
The goal here is to use a combination of clustering techniques
along with Principal Components in order to understanding groupings
of professional basketball players in the National Basketball
Assocation (NBA)


#Load the libraries
```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(ggrepel))
suppressMessages(library(graphics))
suppressMessages(library(factoextra))


```



#Create the data
For our dataset, we will use use Advanced Stats found at 
Set Basketball-reference.com: NBA/Season/Advanced
The data were extracted via some web scraping and data curation
using R's rvest package
Run Rvest_NBA.R to extract the data.


Read the data, it's been saved to file.
```{r}
nba.df0 <- read.csv("NBA2019.csv")


```


How many observations?
```{r}
nrow(nba.df0)

```


Who are the players?
```{r}
(players <- nba.df0$Player)

```


##Some data reorganization
To start, we will subset the data. As a first step,
will just use players who logged the most minutes during the
season. For our purposes, we'll use the top 15%.


Take top % or of players my minutes played
```{r}
(minMins <- with(nba.df0,quantile(MP,0.85)))

```


Only use these players, and LeBron James (who was injured and
wasn't in the top 15% of minutes played)
```{r}
nba.df <- nba.df0 %>% 
    filter(MP>minMins| Player == "LeBron James") 

    
```


##Principal Components
Extract player names and the data fields
```{r}
(players <- nba.df$Player)
nba.df1 <-  nba.df[,8:27]

```


Scale and add row names
```{r}
nba.df1 <- data.frame(scale(nba.df1))
row.names(nba.df1) <- players

```



Here is the Principal Components analysis
```{r}
nba.pca <- prcomp(nba.df1)
```


What does facto tell us?
```{r}
fviz_pca_biplot(nba.pca,repel=TRUE)

```



##Principal Components by hand.
Using the nba.pc data frame, we can extract information and build
custom views.

Pull off the PCA information
This is the change of basis matrix (V in the SVD)
```{r}
rots <- nba.pca$rotation


```


Rotate into the new basis.
```{r}
nba.pca1<- data.matrix(nba.df1) %*% rots
dim(nba.pca1)


```


The first three directions
```{r}
pca1 <- nba.pca1[,1]
pca2 <- nba.pca1[,2]
pca3 <- nba.pca1[,3]



```


##Add in K Means clustering to the Principal Components


Five clusters, this is arbitrary
```{r}
K <- 5
```


run 30 times, allow up to 30  iterations for convergence
```{r}
nba.km <- kmeans(nba.df1,K,iter.max=20,nstart=30)

```


Our clusters
```{r}
nba.km$cluster

```


Add the clusters to the data frame
```{r}
nba.df$cluster.km<- nba.km$cluster
head(nba.df)

```


Look at one cluster
Who are these players...any NBA fans out there?
```{r}
filter(nba.df,cluster.km==2)%>%
    dplyr::select(Player)


```


What are the cluster sizes?
```{r}
with(nba.df,table(cluster.km))

```


Build the rotated data frame
```{r}
nbaCluster.df <- data.frame(pca1,pca2,cluster=factor(nba.df$cluster.km),name=players)

```


plot the cluster.pcas on top of the PCA first two components
```{r}
ggplot(nbaCluster.df,aes(pca1,pca2,color=cluster))+
    geom_point(size=1)+
    geom_text_repel(aes(label=name),size=3)+
    ggtitle("NBA Cluster Analsys and PCA")
```


Anything interesting here?        



#Hierarchical clustering
As an alternative to Kmeans, we can use Hierarchical Clustering
Use R hclust function.

The distance function
```{r}
nba.dist <- dist(nba.df1)
    
nba.hc.c <- hclust(nba.dist,method="complete")
nba.hc.a <- hclust(nba.dist,method="average")
nba.hc.s <- hclust(nba.dist,method="single")


```


Look at the hierarchical plots
Complete
```{r}
par(mfrow=c(1,1))
plot(nba.hc.c,cex=0.5,main="Complete")


```


Average
```{r}
plot(nba.hc.a,cex=0.5)

```


Single
```{r}
plot(nba.hc.s,cex=0.5)

```



Better visuals using ggdendrogram
```{r}
library(ggdendro)

```


Complete
```{r}
ggdendrogram(nba.hc.c,rotate=T)

```


Average
```{r}
ggdendrogram(nba.hc.a,rotate=T)

```


Single
```{r}
ggdendrogram(nba.hc.s,rot=T)

```



We can do better visuals with ggplot
ggdendro library

Extract the gg data using the complete distance
```{r}
nba.dendr <- dendro_data(nba.hc.c)

```



Build the plot
```{r}
nba.dendro.gg <-
    ggplot() +
    geom_segment(data=segment(nba.dendr),
                 aes(x=x, y=y, xend=xend, yend=yend),size=.1) +
    geom_text(data=label(nba.dendr),
              aes(x=x, y=y, label=label, hjust=0),
              size=2,color="blue") +
    coord_flip() +
    scale_y_reverse(expand=c(0.2, 0))+
    theme(axis.line.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.text.y=element_blank(),
          axis.title.y=element_blank(),
          panel.background=element_rect(fill="white"),
          panel.grid=element_blank())+
    guides(color=F)+    
    ggtitle("Hierarchical Clustering of NBA Players")
nba.dendro.gg



```


Identify clusters
Pick a cut level
```{r}
Cut.lev <- 6
cut.labs <- cutree(nba.hc.c,Cut.lev)

```


create data frame to match to player names
```{r}
cut.df <- data.frame(Player=names(cut.labs),
                     cluster=cut.labs)

```



add in cluster information
```{r}
labels.df <- label(nba.dendr)%>%
        inner_join(cut.df,by=c("label"="Player"))

```


Replot with clusters labeled
```{r}
nba.dendro.gg1 <- nba.dendro.gg +
    geom_text(data=labels.df,
              aes(x=x, y=y, label=label,
                  hjust=0,
                  color=factor(cluster)),
              size=2)+
    scale_color_brewer(palette="Set1")+
  ggtitle("Hierarchical Clustering of NBA Players With Grouping")
nba.dendro.gg1


```



Use correlation as the distance metric
A different way of measuring distance is to use 1-cor. This way,
players that are high correlated are close together. 


The tricky bit is to get the right correlation. We could either
look at how the stat variables are correlated, or how the players
are correlated. We want the player correlation.

```{r}
cor.mat <- cor(t(nba.df1))

```


There should be a row/col for each player
```{r}
dim(cor.mat)

nba.dist.cor <- 1-as.dist(cor.mat)
nba.hc.cor <- hclust(nba.dist.cor,method="complete")

nba.dendr.cor <- dendro_data(nba.hc.cor)

```



```{r}
nba.dendro.gg2 <- ggplot() +
  geom_segment(data=segment(nba.dendr.cor),
               aes(x=x, y=y, xend=xend, yend=yend),size=.1) +
  geom_text(data=label(nba.dendr.cor),
            aes(x=x, y=y, label=label, hjust=0), size=2,color="blue") +
  coord_flip() + scale_y_reverse(expand=c(0.2, 0))+
  theme(axis.line.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.y=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_rect(fill="white"),
        panel.grid=element_blank())+
  ggtitle("Hierarchical Clustering using PCA")
nba.dendro.gg2

```



Add cluster cuts

```{r}
Cut.lev <- 6
cut.labs <- cutree(nba.hc.cor,Cut.lev)
cut.df <- data.frame(Player=names(cut.labs),
                     cluster=cut.labs)

```


add in cluster information
```{r}
labels.df <- label(nba.dendr.cor)%>%
        inner_join(cut.df,by=c("label"="Player"))


```


Replot with clusters labeled
```{r}
nba.dendro.gg2 <- nba.dendro.gg2 +
    geom_text(data=labels.df,
              aes(x=x, y=y, label=label,
                  hjust=0,
                  color=factor(cluster)),
              size=2)+
    scale_color_brewer(palette="Set1")+
  ggtitle("Hierarchical Clustering (Correlation)")
nba.dendro.gg2


```



##Heatmaps
This allows us to see clustering in two dimensions
A nice into can be found at: http://www.r-bloggers.com/drawing-heatmaps-in-r/



Heatmap without clustering/
each grid value is the player x stat value (scaled)


Set the color palatte
```{r}
my_palette <- colorRampPalette(c("red", "white", "blue"))(n = 20)
my_palette


```


Build a heat map. This requires a data matrix
```{r}
nba.data <- data.matrix(nba.df1)
heatmap(nba.data,col=my_palette,Rowv=NA,Colv=NA)

```


As displayed, this isn't really informative. We need to arrange the
entries in a natural manner,


Data clustering with trees


Here we have a heatmap with a hierarchical clustering on both the
players and the stat variables
```{r}
heatmap(nba.data,col=my_palette)


```


Use Pearson correlation as the distance function
```{r}
dist.pear <- function(x) as.dist(1-cor(t(x)))


heatmap(nba.data,
        col=my_palette,
        distfun=dist.pear)

```


Intesting...

## Assignment

### Find another data set similar to the NBA dataset
(Baseball Reference is one choice, but there are other, non-sport
options). Perform a complete cluster and PCA analysis. 


The World Bank data is one of the best sources for country level
data. It contains hundreds of variables dating back to the
1970's. The data has been meticulously gatherered and vetted. 
Overall, it's hard to find a better source

You can download data directly from their website at
https://data.worldbank.org/.
There are, fortunately, very nice R interfaces to the data. The
best is wbstats


Load the library
```{r}
library(wbstats)
```




Good introduction to wbstats package as a way of pulling off World
Bank Data
https://cran.r-project.org/web/packages/wbstats/vignettes/Using_the_wbstats_package.html



Start by loading the cached data
```{r}
new_wb_cache <- wbcache()


```


##The wbstats package allows you to search for "indicators" related to any
topic. You just supply a search term. 
Look for indicators related to gini index.
```{r}
gini.ids <- wbsearch("Gini Coefficient")

```


When you are done you have indicator IDs and indicator descriptors
Indicator Ids
```{r}
gini.ids$indicatorID
```


Descriptions
```{r}
gini.ids$indicator




```


The searches can be more  complicated, say using wild cards and
regular expressions.
```{r}
gdp.ids <- wbsearch("gdp.*capita.*US\\$", cache = new_wb_cache)


```


or even more complicated regular expresions. 
```{r}
gdp.ids <- wbsearch("gdp.*capita.*US\\$", cache = new_wb_cache)


```



You can extract all the country names
```{r}
wb_countries <- wbcountries() 
names(wb_countries)
with(wb_countries,table(region))

```


Some of the names are actually regions. It's easy to get rid of
regions. 
```{r}
allCountries <- wb_countries %>%
    filter(region != "Aggregates") %>%
    dplyr::select(country)

```


thesea are the countries
```{r}
allCountries


```



You can extract all the country names
```{r}
wb_countries <- wbcountries() 
names(wb_countries)
with(wb_countries,table(region))

```


Get rid of regions
```{r}
allCountries <- wb_countries %>%
    filter(region != "Aggregates") %>%
    dplyr::select(country)

```


thesea are the countries
```{r}
allCountries


```


##Another method is to go the World Bank website and just explore the data
https://data.worldbank.org/indicator
The indicator name is in the URL selector at the top of the webpage

In any case, here are a few variables we can start workin with. I
picked these by going to the world bank website. 


##Some variables to explore
Population
```{r}
pop <- "SP.POP.TOTL"

```


Population Density
```{r}
popdens <- "EN.POP.DNST"

```


GDP per capital
```{r}
gdp <- "NY.GDP.PCAP.CD"

```


Mortalilty under 5
```{r}
mort <- "SH.DYN.MORT"

```


School Enrollment secondary %
```{r}
secschool <- "SE.SEC.ENRR"

```


agriculltural land
```{r}
ag <- "AG.LND.AGRI.ZS"

```


CO2 Emissions,
```{r}
co2 <- "EN.ATM.CO2E.PC"

```




List the indicators
```{r}
inds <- c(pop,popdens,gdp,mort,secschool,ag,co2)
inds
```


Now we can extract the data from the World Bank databaset. This
takes akes a moment or two.

Grab the data
```{r}
wb_dat <- wb(indicator = inds)


```


Let's see what we have
```{r}
with(wb_dat,table(indicatorID))

```


By date
```{r}
with(wb_dat,table(date,indicatorID))


```


Extract 2014 Data
```{r}
wb2014 <- wb_dat %>%
    filter(date=="2014")


```


Drop regions and other variables
```{r}
wb2014 <- wb2014 %>% 
    filter(country %in% allCountries$country) %>%
    dplyr::select(-iso3c,-iso2c,-date,-indicator)


```


Spread the data so we can use it with clustering
```{r}
wb2014x <- wb2014 %>%
    spread(indicatorID,value)
head(wb2014x)


```


What does the US look like?
```{r}
wb2014x %>%
    dplyr::filter(country=="United States")

```


Define some more suggestive field names    
```{r}
names(wb2014x)[2:10] <-
    c("ag_use","co2","pop_dens","gdp_cap","sec_school","infant_mort","pop")
summary(wb2014x)


```


There are still some NAs. Extract the complete cases. 
```{r}
cc <- complete.cases(wb2014x)
wb.df <- wb2014x[cc,]

```


What's left...
```{r}
nrow(wb.df)


```


Save the data for future use
```{r}
write.csv(wb.df,"worldbank.csv",row.names=F)



```



#Hierarchical Clustering

Grab all the countries
```{r}
(countries <- wb.df[,1])

```


How many countries do we have?
```{r}
(numCountries <- nrow(wb.df))

```


Scale the data. 
```{r}
wb.scale <- scale(wb.df[,-1])

```


Assign row names, this aids in plotting
```{r}
row.names(wb.scale) <- countries




```


Compute the pair-wise distances
This computes aall
```{r}
wb.dist <- dist(wb.scale)
str(wb.dist)

```


Note: We have this choose(numCountries,2) pairs of values
```{r}
numCountries*(numCountries-1)/2



```


##Hierarchical Clustering
```{r}
par(mfrow=c(1,1))
hc1 <- hclust(wb.dist,method="complete")
plot(hc1,cex=.5)

```


Look at all four clustering methods
Note how different the shapes are
```{r}
op <- par(mfrow=c(2,2))
hc1 <- hclust(d1,method="complete")
plot(hc1,cex=.5)

hc2 <- hclust(d1,method="average")
plot(hc2,cex=.5)


hc3 <- hclust(d1,method="single")
plot(hc3,cex=.5)



hc4 <- hclust(d1,method="centroid")
plot(hc4,cex=.5)

par(op)


```


Assignment: World Bank Data
Extract a more interesting set of parameters. Explore the World
Bank site and get at least 12 different (more is ok)
features. Select at least two different years (separated in
time). How do the clusters looks? If you add Principal Components,
do you see any natural clustering behavior relative to the
Principal Components?